\chapter{Data Analysis}
\label{data-analysis}

\section{Descriptive Statistics}

Point estimate of a population proportion. Consider the
  pizza delivery data that is available on the class Moodle
  page. Construct a point estimate of the probability that the amount
  of tips received in a shift is greater than \$60. What is the
  standard error of your point estimate? You can do the calculations
  by hand or use the software of your choice. If you use software, you
  can use it in any way you like.  For example, I used R as a
  calculator to simply help with the required computations.
  
\begin{Verbatim}
pizza <- read.table("pizza.txt", header=TRUE)
attach(pizza)
x <- sum(Tips > 60)
n <- length(Tips)

# follow the formula for the point estimate and the standard
# error of a sample proportion
phat <- x/n
se <- sqrt((phat*(1-phat))/n)

> phat
[1] 0.2413793
> se
[1] 0.0212373
\end{Verbatim}

% discussion of confidence interval.
% change problem context, have >= 30 observations, use Normal distribution.
  \emph{Confidence Interval.}  An article in the \emph{Journal of Heat
    Transfer} describes a method of measuring the thermal conductivity
  of high-purity iron.  Using a temperature of $100^{\circ}$F and a
  power input of 550 W. The following 10 measurements of thermal
  conductivity (in Btu/hr -- ft -- ${}^{\circ}$F) were determined.
\begin{align*}
  41.60,~41.48,~&42.34,~41.95,~41.86\\
  42.18,~41.72,~&42.26,~41.81,~42.04
\end{align*}
A point estimate of the population mean thermal conductivity
(at $100^{\circ}$F and 550 W) is the sample mean
\[
  \mean{X} = 41.924
\]
The standard error of the sample mean (i.e. the standard error
of the point estimate) is
\[
  \text{se}(\mean{X}) = \frac{S}{\sqrt{n}} = \frac{0.284}{\sqrt{10}} = 0.0898
\]
where $S$ is the sample standard deviation. Notice that the standard
error is about 0.2 percent of the sample mean, indicating a relatively
precise point estimate of thermal conductivity. Occasionally you will
hear people refer to the coefficient of variation (CV).
\[
  \text{CV} = \frac{S}{\mean{X}}
\]
The CV is another measure of the spread of the data. \emph{Question:}
What are the units of the CV?

Now suppose we want to construct a 95\% confidence interval for the
population mean thermal conductivity $\mu$. So, our confidence level
will be $1-\alpha=.95$, and $\alpha=.05$. Since there are only 10
sample data points, we will use the $t$ distribution. From our
discussion in class we know that a $(1-\alpha)$\% confidence interval
for $\mu$ is
\[ \mean{X} \pm t_{\alpha/2,n-1}\frac{S}{\sqrt{n}} \]
From the tabulated values of the $t$ distribution, we see that
$t_{\alpha/2,n-1} = t_{.025,9} = 2.262$. A 95\% confidence interval for $\mu$ is
\[  41.924 \pm 2.263 \times \frac{0.284}{\sqrt{10}} \]
or (41.721, 42.127).

Using R, we could do the following

\vspace{.1in}
\begin{Verbatim}[samepage=true]
> x <- c(41.60,41.48,42.34,41.95,41.86,42.18,41.72,42.26,41.81,42.04)
> alpha <- .05
> n <- length(x)
> xbar <- mean(x)
> se <- sd(x)/sqrt(n)
> cp <- qt(1-alpha/2,n-1)

> xbar + c(-1,1)*cp*se
[1] 41.72076 42.12724
\end{Verbatim}
\vspace{.1in}
To use the Normal distribution instead of the $t$ distribution, obtain the
critical point as
\begin{Verbatim}
> cp <- qnorm(1-alpha/2)
\end{Verbatim}

\emph{Question:} When I got the critical point in R, why did I use
\texttt{qnorm(1-alpha/2)} and not \texttt{qnorm(alpha/2)}?

\section{Descriptive Graphics}

\section{Exercises}

\begin{enumerate}
\subsubsection*{Descriptive Statistics}

% need to use different data values
% solution in ie1101/hwfa2018/hw3
\item \emph{Taking averages.} Consider two datasets: 1, 5, 9 and 2,
    4, 6, 8.
\begin{enumerate}
\item Denote the sample means of the two datasets by $\mean{x}$ and
      $\mean{y}$. Is it true that the average
      $\left(\mean{x}+\mean{y}\right)/2$ of $\mean{x}$ and $\mean{y}$ is
      equal to the sample mean of the combined dataset with 7
      elements?
\item Suppose we have two other datasets: one of size $n$ with sample
  mean $\mean{x}_n$ and another dataset of size $m$ with sample mean
  $\mean{y}_m$. Is it always true that the average
  $\left(\mean{x}_n + \mean{y}_m\right)/2$ of $\mean{x}_n$ and
  $\mean{y}_m$ is equal to the sample mean of the combined dataset with
  $n + m$ elements? If no, then provide a counterexample. If yes, then
  explain this.
\item If $m=n$, is $\left(\mean{x}_n + \mean{y}_m\right)/2$ equal to the
  sample mean of the combined dataset with $n + m$ elements?
\end{enumerate}

% this problem is OK.
% solution in ie1101/hwfa2018/hw3
\item \emph{Computing the sample variance.} The following rule is
  useful for the computation of the sample variance (and standard
  deviation). Show that

  \[ \frac{1}{n}\sum_{i=1}^n(x_i - \mean{x}_n)^2 = \left(\frac{1}{n}\sum_{i=1}^nx_i^2 \right)- \left(\mean{x}_n\right)^2 \]
    where $\mean{x}_n = \left(\sum_{i=1}^n x_i\right)/n$

% re-written by Braeden
\item \emph{Summarizing a data set with statistics.}  In the
  \texttt{datasets} package in R, there is a data set called
  \texttt{discoveries} which contains the numbers of "great"
  inventions and scientific discoveries in each year from 1860 to
  1959.

\begin{enumerate}
\item From the data set of 100 observations, determine the following
  summary statistics: minimum, maximum, mean, median, mode, first
  quartile, third quartile, and standard deviation.
\item Create a table that shows the distinct values of the number of
  discoveries per year and their counts, i.e. the number of times that
  each value occurred.
\end{enumerate}

% re-written
\item \emph{Point estimate of a population mean.} Suppose the
  following data points are a sample of a golfer's scores over his
  last 20 rounds. Construct a point estimate of his average
  score. What is the standard error of your point estimate?
\begin{verbatim}
73,69,65,70,67,67,78,72,74,71,70,69,70,67,68,73,70,77,72,69
\end{verbatim}
  
% re-written
\item \emph{Standard error when estimating a proportion.} In a survey, a
  random sample of \num{1200} students are asked whether they prefer
  online or in-person classes.  Out of the \num{1200} students,
  \num{424} said they prefer online classes. Compute a point estimate
  of the overall proportion of students that prefer online classes and
  calculate the standard error of your estimate.

% this problem is OK
\item \emph{Lognormal distribution parameter estimation.}  The file
  \texttt{component-lifetimes.txt} contains the time to failure for
  \num{1345} components (in hours). The times are known to come from a
  Lognormal distribution. 
\begin{enumerate}
\item Estimate the parameters of the failure time distribution. 
\item Use the parameters to estimate the mean time to failure. 
\item Use the parameters to estimate the probability that a component lasts
  longer than \num{10000} hours.
\end{enumerate}

% tag-and-recapture example? but not fish, maybe a wolf or lynx population in MN?
\item \emph{Estimation of the size of a population.}


% this problem is OK
\item \emph{Confidence interval for a mean.} Restaurants are making
  more use of their data on service times for planning purposes.  The
  data in the file \texttt{restaurant-service-times.txt} contains 220
  observations on the time in minutes from seating until departure in
  one particular restaurant. 
\begin{enumerate}
\item Construct a 95\% confidence interval for the true mean service
  time. Remember that this data is just a sample from a larger
  population, the true distribution of which is unknown.
\item Do you think that the Central Limit Theorem applies to this
  data? Why or why not?
\end{enumerate}

% re-written by Braeden
\item \emph{Validation of a simulation model.} A simulation model of a
  job shop was developed to investigate different scheduling rules. To
  validate the model, the scheduling rule currently used was
  incorporated into the model and the resulting output was compared
  against observed system behavior. By searching the previous year’s
  records, it was estimated that the average number of jobs in the
  shop was 22.5 on a given day. The file \texttt{simulation.txt} contains the 
  results for the average number of (simulated) jobs in the shop for 30 independent 
  replications of the simulation model, each for 30 days of simulated time.

  One metric that can be used to validate the simulation model is to
  construct a 95\% confidence interval for the true mean number of
  (simulated) jobs in the shop. If the confidence interval
  contains the value 22.5, then, the model captures the average
  work-in-process for the job shop\footnote{You would want to know that the
  baseline simulation model is accurate before simulating any proposed
  changes to the scheduling rules.}. Construct the confidence interval
  and comment on the result.

\item \emph{Confidence interval for a proportion.} 

\item \emph{Comparison of system configurations.}


% this problem needs a different data set
\item \emph{Bootstrap confidence interval.} The data file
  \texttt{clinical-trial.txt} contains data on 20 patients. 10
  patients were randomly assigned to receive Medicine A, and 10 were
  randomly assigned to receive Medicine B. The data represents the
  responses of the patients to their assigned medicines. Use the
  bootstrap technique to determine whether or not there is a
  difference in the median response between the two medicines.  In
  order to do this, you should take $B$ bootstrap samples of the data
  (where $B \geq 200$). For each bootstrap sample, compute the
  difference in median response between the two medicines. That is to
  say, for each bootstrap $i$ sample you will compute
  \[
  \text{median}(A_i) - \text{median}(B_i)
  \]
  where $\text{median}(A_i)$ is the median of the data associated with
  medicine A in the $i$th bootstrap sample ($i$ goes from 1 to
  $B$). Determine a 95\% confidence interval for the difference in
  median response by taking the .025 and .975 quantiles of the
  bootstrap replicates. If your confidence interval does not contain
  zero, then you should conclude that there is a difference. If that
  is the case, then state the direction of the difference. That is,
  state which medicine has a higher median response. \label{ex:boot}


% this problem is OK
\item \emph{Required sample size.} The following data are observations
  from a past study of hummingbird migration rates in miles flown per
  day. These observations are from 30 different birds. A researcher
  would like to construct a two-sided, 95\% confidence interval for
  the average rate (in miles per day). The researcher would like for
  the width of the confidence interval to be no larger than 1 day. It
  is very expensive to attach identifiers to the birds, and so the
  researcher has asked you to determine the smallest sample size that
  will achieve the desired confidence interval. What sample size do
  you suggest?
\begin{Verbatim}[samepage=true]
17,17,22,18,19,21,21,23,21,25,
19,21,19,20,20,21,19,20,18,17,
18,20,19,23,18,22,18,22,18,24
\end{Verbatim}

\begin{solution}
\bs The width $W$ of a confidence interval is
2 times the half-width
\[
W = 2 \times z_{\alpha/2} \times \frac{S}{\sqrt{n}}
\]
and the researcher would like $W\leq 1$. We can use the data to
estimate a standard deviation $S = 2.133$. We do not know $n$. In
fact, that is the question we are trying to answer. You can assume
that you will have a sample size at least as big as, say, 30
birds. Using $n=30$ and the tabulated values for the Normal
distribution, we find that $z_{\alpha/2} = z_{.025} = 1.96$.  Then,
solving for $n$, the required sample size is
\[
n \geq 4 \times (z_{\alpha/2} \times S)^2 = 
4 \times (1.96 \times 2.133)^2 = 69.9
\]
We would recommend a sample size no smaller than 70 observations.
Note that you will want to use the 30 observations that you already
have, so a confidence interval of width 1 day will require 
approximately 40 additional observations.
\end{solution}

\subsubsection*{Descriptive Graphics}

% this problem is OK.
\item \emph{College students and driving speed.} The file
  \texttt{speed\_gender\_height.csv} contains 1,325 observations on
  gender, height, and the fastest speed ever driven (in mph) for a
  sample of college students.

\begin{enumerate}
\item Create a boxplot of speed by gender. That is to say, make one
  boxplot for males and one boxplot for females, but put them
  side-by-side on the same plot.
\item Make an x--y plot with height on the x--axis and speed on the
  y--axis. Color the plotted points according to gender. Place a
  legend that shows the color associations. Another option is to use
  different plotting symbols rather than color to distinguish males
  and females.
\end{enumerate}

\end{enumerate}
