\chapter{Predictive Modeling}

\section{Linear Regression}

\section{Time Series}

\section{Exercises}

\begin{enumerate}

% this problem is OK
\item \emph{Linear regression with a single predictor variable.}
In the \texttt{datasets} package in R there is a data set named
\text{faithful} that
contains data on the Old Faithful geyser in Yellowstone National Park,
Wyoming, USA.  The variables in the data set are
\begin{compactitem}[$\circ$]
\item \texttt{eruptions} the eruption time in minutes
\item \texttt{waiting} the time in minutes until the next eruption
\end{compactitem}

To view information about the data set, type
\begin{Verbatim}
> ?faithful
\end{Verbatim}
at the R prompt. Use this data to perform the following exercises.

\begin{enumerate}
\item Create histograms of \texttt{eruptions} and \texttt{waiting}
  (separately). \label{hist}

\item Create a scatter plot (i.e. an x--y plot) with \texttt{eruptions}
  on the x axis and \texttt{waiting} on the y axis. \label{scat}

\item From the histograms and the scatter plot that you created, what
  can you say about the behavior of the Old Faithful geyser? \label{interp}
  
\item Fit a linear regression model (using the \texttt{lm()} function
  with \texttt{waiting} as the response variable and \texttt{eruptions}
  as the only predictor variable. Print a summary of the results.
  \begin{enumerate}
  \item What is the interpretation of the intercept? \label{int}
  \item What is your interpretation of the fitted coefficient
    on \texttt{eruptions}? \label{dur}

  \item You just observed an eruption of duration 4 minutes.
    Make a prediction on how long you will have to wait until
    the next eruption. Can you make any statement about the
    uncertainty in your prediction? In other words, can
    you give a range for the time until the next eruption?
    Don't worry about being exact with your range, just
    give something reasonable. \label{pred}.
  \end{enumerate}
\end{enumerate}


% this problem needs to be re-written
\item \emph{Using linear regression to summarize a dataset.}
  Download the wine data set \texttt{Wine.csv} from Canvas. Fit a
  linear regression model using \texttt{lm()} with a price index
  (\texttt{Price}) as the response variable and the following
  predictor variables:
  \begin{compactitem}
  \item growing season temperature in $^{\circ}$C (\texttt{AGST}),
  \item winter rain amount in ml (\texttt{WinterRain}),
  \item harvest rain amount in ml (\texttt{HarvestRain}), and
    \item age of vintage in years (\texttt{Age}).
    \end{compactitem}
    So the model is
    \[ Price_i = \beta_0 + \beta_1 AGST_i + \beta_2 WinterRain_i +
      \beta_3 HarvestRain_i + \beta_4 Age_i + \epsilon_i \]
    Now do the following.
    \begin{enumerate}
    \item Looking at the summary of the fitted model, the intercept
      $\beta_0 \approx -3.43$. What is the interpretation of $\beta_0$?
      In other words, I don't think that the price of wine can be negative,
      so how should we interpret the intercept?
  \item Plot \texttt{Price} as a function of
    \texttt{HarvestRain}. Overlay a fitted regression line from the
    full model onto the plot.  When plotting the regression line you
    should show \texttt{Price} at the average \texttt{AGST}, average
    \texttt{WinterRain}, and average \texttt{Age}.
    In other words, it's a two-dimensional plot, but for the other
    variables that are not shown, we compute \texttt{Price} at their
    average values. So you want to overlay
    \[ Price_i = \beta_0 + \beta_1 \overline{AGST} +
      \beta_2 \overline{WinterRain} +
      \beta_3 HarvestRain_i +
      \beta_4 \overline{Age} \]
    onto the data. You can use \texttt{coef()} to extract the
    coefficients from the fitted model object.

  \item If you are a wine-maker, when would you like to see
    rain?

  \item Plot the actual \texttt{Price} vs. the predicted (fitted)
    price. If your fitted model is stored in an object named
    \texttt{fm}, then you can get the predicted price as follows.
    \begin{Verbatim}
      wine$pred <- fitted(fm)
    \end{Verbatim}
    % $
    or
    \begin{Verbatim}
      wine$pred <- predict(fm)
    \end{Verbatim}
    % $

  \item In the summary output of the fitted model, the estimated residual
    standard error is reported to be
    $\hat{\sigma}_{\epsilon}=0.295$. Independently compute this quantity. In
    other words, use the actual values from the data and the fitted
    values from the model to compute the residual standard error
    yourself.  The formula is
    \[ \hat{\sigma}_{\epsilon} = \sqrt{ \frac{\sum_{i=1}^n \left(y_i - \hat{y_i}\right)^2}{n-k}} \]
    where $y_i$ and $\hat{y_i}$ are the actual and fitted values of observation
    $i$, respectively, $n$ is the total number of observations, and $k$ is the
    number of fitted parameters in the model. $n-k$ is the degrees of freedom.

  \item Do you think that a linear model is appropriate for this data?
  \end{enumerate}
    
\end{enumerate}
